Comparing R packages for calculation of persistent homology

### 2020-01-18 tdabench-paper/tdabench.Rmd
Imported all other text from Overleaf, Revised Intro

\section{Introduction}

\section{Mathematics of Persistent Homology}

\begin{figure}
  \centering
  \includegraphics[height=2.5in]{../Figures/Final_Figures/Intro_Rips.png}
  \caption{As $\delta$ continues to grow, more circles intersect filling out the simplicial complex. Features on the simplicial complex appear and die as $\delta$ increases. Eventually, the full convex hull is drawn ending the "filtration" process.}
\end{figure}
\begin{figure}
  \centering
  \includegraphics[height=2.5in]{../Figures/Final_Figures/Intro_Alpha.png}
  \caption{As $\alpha$ continues to grow, more circles fill out the Voronoi cells. Once two circles in adjacent Voronoi cells fill out the cell, a connecting segment can be drawn. Eventually, once $\alpha$ is large enough, the Delaunay Triangulation is formed.}
\end{figure}

>>>>

\section{Introduction}
Topological data analysis (TDA) is a broad set of methodologies that characterize structural features of data sets inspired by topological principles.
Within the umbrella of TDA, persistent homology represents an algebraic approach to understanding the number, characteristics, and persistence of structural features in an n-dimensional point cloud.

In the basic workflow of persistent homology, a series of simplicial complexes are generated on point clouds to characterize the topological features. 
Simplicial complexes are connected sets of points, lines, and n-dimensional shapes.
There are several methods to generate these complexes on point clouds. 
In this paper, we are using the Rips and Alpha Complex for our persistent homology calculations. 
In both methods, a series of data structures, called simplicial complexes, are built on top of a point cloud.
These complexes can be viewed as approximating real topologic relationships in the data.
The exact method of constructing these complexes is described in the Mathematics section. 
However, the key takeaway is that the simplicial complex attempts to reveal the underlying topology of the point cloud. 
These features can be described by dimension and "persistence." 
Long lasting (persistent) features that exist across the series of simplicial complexes intuitively represent "real" geometrical patterns as opposed to noise.

One limitation of persistent homology is its run time limits for processing large numbers of points ( $> 10^9$ points) and calculating of high dimensional features ($>$ 2 dimensions). 
Alpha complex calculations have a complexity of $O(n^\frac{d}{2})$ and Rips complex calculations have a complexity of $O(2^n)$ where n is the number of points and d is the dimension of the point cloud \cite{roadmap}. The limits and computational resources needed to run complex data sets with major persistent homology C++ software have been well characterized \cite{roadmap}.

There are several C++ libraries that calculate Alpha and Rips complexes.
Dionysus provides Alpha and Rips complex functionality \cite{Dionysus}.
GUDHI is another C++ library that also provides both functionalities as well as other TDA methodologies \cite{gudhi}. These libraries have been implemented in R with the TDA package \cite{TDA} providing an easier interface to utilize persistent homology calculations.

A new C++ library, Ripser, was developed as a more efficient way to calculate Rips Complexes. The code is supposed to outperform other similar libraries (e.g. Dionysus and GUDHI) by a factor over 40 in runtime \cite{Ripser}. 
This library was given an R interface in the package TDAstats \cite{TDAstats}. 
TDAstats represents a more specialized package for topological data analysis compared to the TDA package.
It calculates persistent homology using the more efficient Ripser algorithm and provides data visualization with integration of ggplot. While Ripser is known to be faster than the other libraries for Rips complex calculations in C++, how this translates in the R interface is not clear.
While C++ generally runs faster than R, R provides higher level functionality, and appeals to a broad audience of data science users.

Topological data analysis has had a broad range of usage from viral evolution to physical chemistry~\cite{TDA-Viral,TDA-PChem}. 
Given that persistent homology can apply to any n-dimensional data set, it has wide ranging applicability. 
Its main limiting factor is computational for calculating high dimensional features in large multi-dimensional data sets. 
We believe lack of analysis of these higher dimensional features may under utilize persistent homology's analysis potential.
In this project, we explore different methodologies to find efficient options that could solve this issue. 
We specifically aim to characterize the computational performance (run time) and limits of persistent homology calculations between Rips complexes in the R packages TDAstats and TDA.
We also explore the differences in computational performance between Rips and Alpha complexes in R to assess which datasets are most appropriate for each algorithm.

\section{Mathematics of Persistent Homology}
This section summarizes the mathematics behind generating simplicial complexes on point clouds in the workflow of persistent homology. 
An n-dimensional simplex is the convex hull of n+1 vertices.
More intuitively, n-dimensional simplex is the simplest n-dimensional object (e.g. a 1 simplex is a line, a 2-simplex is a triangle, 3-simplex is a tetrahedron, etc.).
These simplices can be glued together on a common subsimplices to form a simplicial complex (e.g. two triangles sharing a common side).
In a simplicial complex, certain features will arise that can be characterized by dimension number (Betti number): 0-dimensional features refer to connected components, 1-dimensional features refer to holes, 2-dimensional features refer to cavities, etc. ~\cite{phom-survey}

There are several different methods to construct a simplicial complex on a given point cloud S, but this paper focuses on the Rips complex and Alpha complex. 
The Rips complex is perhaps the most common method for constructing a simplicial complex under persistent homology ~\cite{Rips-Complex}.
In a point cloud of k points, a distance parameter, $\delta$, can be used to draw a circle of diameter $\delta$ around every single point in S.
If the distance parameter is non-zero, then some of the resulting circles may intersect.
If they intersect, a line is drawn to connect the points (which represent the centers of the growing circles).
This group of points and lines form a simplicial complex.
For each distance parameter, $\delta$, there will be some simplicial complex associated with it.
As $\delta$ increases, different topological features may appear, persist, then disappear once $\delta$ has reached a threshold.
Eventually, once $\delta$ reaches the maximum euclidean distance between any pair of points in the point cloud, a convex hull will form around all k points creating a (k-1)-dimensional simplex.
By this stage, all features should have disappeared.
A three column matrix can be created recording the Betti number of a certain feature, the distance, $\delta$, at which that feature appeared, and the $\delta$ at which that feature disappeared.
This matrix characterizes the persistent homology of that point cloud.

\begin{figure}
  \centering
  \includegraphics[height=2.5in]{../Figures/Final_Figures/Intro_Rips.png}
  \caption{As $\delta$ continues to grow, more circles intersect filling out the simplicial complex. Features on the simplicial complex appear and die as $\delta$ increases. Eventually, the full convex hull is drawn ending the "filtration" process.}
\end{figure}

Alpha complexes provide another method to generate simplicial complexes on the point cloud S.
Alpha complexes are closely linked with Voronoi diagrams and Delaunay triangulations.
The Voronoi diagram gives each point in S a cell such that each every point in the cell is closest to that point in S.
The nerve of a Voronoi diagram is equivalent to the Delaunay Triangulation~\cite{alpha-complex}.
Alpha complexes are simplicial complexes that are subsets of the Delaunay Triangulation.
The parameter, $\alpha$, can describe the radius of a ball of each point in the point cloud S much like $\delta$ describes the diameter of a circle in the Rips Complex.
An alpha complex at radius $\alpha$ is formed through connecting centers of intersecting regions formed by the intersection of $\alpha$ radius balls and the Voronoi diagrams. 
Eventually, once $\alpha$ is large enough, the full Delaunay complex is formed.
In between these stages, the birth and death of features at certain values of $\alpha$ can be captured in a three column persistent homology matrix much like the Rips Complex.

In both methods, the boundary matrix records all simplicial complexes for a given parameter value. 
From this matrix, the persistent homology matrix is calculated. 
The size of the boundary matrix can describe the memory complexity on the RAM for persistent homology calculations. 
We compare this memory complexity between Alpha and Rips complexes for this project.

\begin{figure}
  \centering
  \includegraphics[height=2.5in]{../Figures/Final_Figures/Intro_Alpha.png}
  \caption{As $\alpha$ continues to grow, more circles fill out the Voronoi cells. Once two circles in adjacent Voronoi cells fill out the cell, a connecting segment can be drawn. Eventually, once $\alpha$ is large enough, the Delaunay Triangulation is formed.}
\end{figure}
